\demes: a pain-free specification for complex demography

### 2021-06-15 paper.tex
First pass at introduction.

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\demes}[0]{\texttt{demes}}
\newcommand{\Demes}[0]{\texttt{Demes}}
\newcommand{\YAML}[0]{\texttt{YAML}}
\newcommand{\moments}[0]{\texttt{moments}}
\newcommand{\dadi}[0]{\texttt{$\partial$a$\partial$i}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\slim}[0]{\texttt{SLiM}}
\newcommand{\gadma}[0]{\texttt{GADMA}}

@@

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to infer complex
demography and study life history at fine scales. An integral component to such
population genetics studies is simulation. Software to either simulate whole
genome sequences
\citep{thornton2014c++,kelleher2016efficient,haller2019slim,adrion2020community}
or informative summary statistics of diversity
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring} have
enabled the increasing complexity of genomic studies, with many software
packages able to handle large sample sizes, many interacting populations, and
deviations from random-mating models of panmixia.

The ability to simulate complex demographic scenarios, however, does not come
without its own set of obstacles and frustrations. First, specifying models
with many populations and parameters can be cumbersome and error-prone (e.g.,
\citep{ragsdale2020lessons}). It is time-consuming and tedious to implement and
then verify the correctness of such demographic models. Making matters worse,
each software package typically has its own application programming interface
(API), which makes it difficult to translate models between simulation
software.

Second, many software API are not particularly easy to read, especially when
there are a large number of demographic events and parameters. This not only
makes debugging difficult, but it can be difficult to even notice if a mistake
has been made. A common interface for unambiguously specifying demographic
models would reduce implementation errors and promote ease-of-use for many
simulation software packages.

Finally, simulated or inferred demographic models are regularly shared in the
literature, and for someone else to be able to reproduce or reimplement some
demographic scenario, it needs to be described fully and without ambiguity.
However, it is common for reported demographic models to lack clear or even
complete descriptions, hindering reproducibility. A human-readable description
for such models would facilitate their sharing and reimplementation.

Here, we present \demes, a specification of complex demographic models that is
both human readable and can be directly passed to a growing number of
simulation software. \Demes\ supports model specification in a high-level
\YAML\ format \citep{ben2009yaml} that is designed to be as easy as possible to
implement, read, and share. Models are then represented in an unambiguous
low-level format within Python that checks model validity. The initial Python
release of \demes\ (version 0.1, described here) supports \emph{static}
demographic descriptions, that is, models with fixed parameters. Future
releases will support [[drawing parameters from distributions, for example for
ABC-based inference, or describing ``dynamic`` models where parameters may be
fit to data, e.g. \moments\ or \dadi]].

\section*{The \demes\ specification}

@@

\subsection*{High-level model specification in \YAML}

@@

\fwdpy &
    Forward-time Wright-Fisher simulation \citep{thornton2014c++}\\

@@

\bottomrule
\end{tabular}
\end{center}
\caption{Software support for Demes. We have included software infrastructure developed
for working with Demes models (parsing, validation, visualisation, etc)
as well as all known downstream software that implements the specification,
at the time of writing.}
\end{table}

\subsection*{Other stuff...}

\begin{itemize}
    \item General framework for inference (have been working on that over in \moments,
        and perhaps it could be generalized to work with any simulation engine
    \item Defining distributions of parameter values, perhaps?
\end{itemize}

>>>>

% local definitions
\newcommand{\ms}[0]{\texttt{ms}}
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
\newcommand{\demes}[0]{\texttt{demes}}
\newcommand{\Demes}[0]{\texttt{Demes}}
\newcommand{\moments}[0]{\texttt{moments}}
\newcommand{\dadi}[0]{\texttt{$\partial$a$\partial$i}}
\newcommand{\fwdpy}[0]{\texttt{fwdpy11}}
\newcommand{\slim}[0]{\texttt{SLiM}}
\newcommand{\gadma}[0]{\texttt{GADMA}}

@@

The ever-increasing amount of genetic sequencing data from genetically and
geographically diverse species and populations has allowed us to infer complex
demography and study life history at fine scales. An integral component to such
population genetics studies is simulation. Software to either simulate whole
genome sequences
\citep{thornton2014cpp,staab2015scrm,kelleher2016efficient,haller2019slim}
or informative summary statistics of diversity
\citep{gutenkunst2009inferring,kamm2017efficient,jouganous2017inferring} have
enabled the increasing complexity of genomic studies, with several software
packages capable of working with large sample sizes, many interacting populations, and
deviations from random-mating models of panmixia.
This ability to infer and simulate such complex demographic scenarios, however,
has highlighted a major shortcoming in community standards. The fragmented
landscape of different approaches to describing demographic models makes
it difficult to compare inferences made by different methods,
and to reliably simulate from previously inferred models.

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
% Thus, any comparison
% between inferences requires these file formats to be parsed, with the user
% being left with the task of unifying the data models.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately there is often ambiguity in these descriptions and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
The \stdpopsim\ project is a community-maintained catalog of species
information and demographic models designed to provide standardised
population genetic simulations~\citep{adrion2020community}.
The development process for adding demographic models to
\stdpopsim\ tries to avoid errors in model implementation by requiring two
independent implementations of each model
to be equivalent. While this process is
robust and has uncovered both significant
errors~\citep{ragsdale2020lessons} and ambiguities in
published models, it is labour intensive.
It would be far better if published results
were reported in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
large number of different methods have been developed
over the past three decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approches to specifying
the demographic model to simulate,
using either a command line
% JK: is there a simulator that just uses a CLI that's not using ms format?
interface~\citep[e.g.][]{hudson2002generating},
a custom input file format~\citep[e.g.][]{excoffier2011fastsimcoal,guillaume2006nemo},
or an Application Programming Interface (API)~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,staab2015scrm},
and the POPDemog visualisation method~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and (potentially) also allow for documentation in the
form of comments. However, there is no standard format shared across
simulators, and the details of the demographic models are not separated
from other simulation parameters. The APIs use by individual simulators
to define demographic models are also not appropriate as an interchange
standard because of the dependency on both the programming language involved,
and the specific details of the simulator in question.

Here we present ``Demes'', a data model and file format
specification for complex demographic models developed by the PopSim
Consortium. The Demes data model precisely defines key parameters
of populations and their relationships over time in an extensible way,
and provides a way to explicitly encode all information relevant to
demography while avoiding repetition. This data model is implemented
in the widely used YAML format~\citep{ben2009yaml}, which provides a
good balance between human and machine readability.
The specification precisely
defines the required behaviour of implementations, ensuring that there
is no ambiguity of interpretation, and includes both a reference
implementation and an extensive suite of test examples and their expected
output. The initial software ecosystem includes high-quality
Python and C parser implementations, as well as utilities for verification
and visualisation of Demes models, and has been implemented in
several popular inference and simulation methods
(Table~\ref{tab-software}).
We hope that this
data model and file format will be widely adopted by the community,
such that users can expect to simulate directly from
inferred models with little or no programming effort.

\section*{Demes}

@@

\subsection*{High-level model specification in YAML}

@@

\fwdpy &
    Forward-time Wright-Fisher simulation \citep{thornton2014cpp}\\

@@

\bottomrule
\end{tabular}
\end{center}
\caption{\label{tab-software}
Software support for Demes. We have included software infrastructure developed
for working with Demes models (parsing, validation, visualisation, etc)
as well as all known downstream software that implements the specification,
at the time of writing.}
\end{table}

\section*{Discussion}

This is an unstructured list of things we should mention:

\begin{itemize}
    \item Recent developments in software engineering have favoured
    static, declarative configuration formats. It feels like a good idea to
    have a Turing complete configuration format, but it turns out not to be.

    \item Drawing parameters from distributions. Why we are static.
    % https://github.com/popsim-consortium/demes-paper/issues/3

    \item Why we've avoided turning this into a model of everything, allowing
    for the specification of genome properties like recombination rates,
    etc.

    \item
    % JK: I thought there was another simulation frontend somewhere..
    Previous efforts such as \texttt{coala}~\citep{staab2016coala} have focused on
    abstracting the differences between individual simulators by providing
    a frontend.

    \item
    % https://github.com/popsim-consortium/demes-paper/issues/11
    We should mention coasim~\citep{mailund2005coasim} because its model
    seems quite close to what we have. It's not suitable as a standard though
    because it uses a programming language.

\end{itemize}
